### 2019-11-26 ic-journal/coarsest.tex
text

In this section we give a proof of the second part of
Theorem~\ref{t:rapproxCongruence}, i.e. $\rapprox$ is the coarsest
congruence contained in $\wb$. This theorem is of special \hl{interest} to
us, because under our \hl{CCS syntax (with only binary summations)} it seems impossible to prove it
without any \hl{antecedent such as ``the involved processes do not
  use up all available labels''}. The general form of this theorem can be
expressed informally as below~\cite{van2005characterisation,Gorrieri:2015jt,Mil89}:
\begin{proposition}
\label{prop:coarsest}
  \hl{Rooted bisimilarity ($\rapprox$) is the coarsest congruence
    contained in weak bisimilarity ($\wb$):}
  \begin{equation}
    \label{eq:coarsest}
\forall p\ \ q.\ p\ \rapprox\ \! q\ \Longleftrightarrow\ ( \forall r.\ p\ +\
r\ \approx\ q\ +\ r )\enspace.
\end{equation}
\end{proposition}
From left to right (\ref{eq:coarsest}) trivially holds, due to the substitutivity of
$\rapprox$ \hl{for} summation and \hl{the fact that} $\rapprox$ implies $\wb$:
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{p} \HOLBoundVar{q}. \HOLBoundVar{p} \HOLSymConst{\HOLTokenObsCongr} \HOLBoundVar{q} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLBoundVar{p} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{q} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r}\hfill{[COARSEST_CONGR_LR]}
\end{alltt}
Thus we are only interested in (\ref{eq:coarsest}) from right to left.
%
The standard argument~\cite{Mil89} requires that $p$
and $q$ do not use up \hl{all} available labels (i.e.~visible actions).
Formalising this \hl{result, however,} requires a detailed treatment on free and bound names of CCS
processes (with the restriction operator being a binder).
\hl{In} this project we do not discuss free and bound names (also
called ``sorts'') of CCS processes. Instead, we
assume \hl{a weaker condition} that all immediate \emph{weak} derivatives of
 $p$ and $q$ do not use all available \hl{labels}. We \hl{might} call it
 the \emph{free action} property:
\begin{alltt}
   \HOLConst{free_action} \HOLFreeVar{p} \HOLTokenDefEquality{} \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{a}. \HOLSymConst{\HOLTokenForall{}}\ensuremath{\HOLBoundVar{p}\sp{\prime}}. \HOLSymConst{\HOLTokenNeg{}}\ensuremath{(}\HOLFreeVar{p} \HOLTokenWeakTransBegin\HOLConst{label} \HOLBoundVar{a}\HOLTokenWeakTransEnd \ensuremath{\HOLBoundVar{p}\sp{\prime}}\ensuremath{)}\hfill{[free_action_def]}
\end{alltt}

Below we show how \hl{Prop.}~\ref{prop:coarsest} is connected with
the statement in Theorem~\ref{t:rapproxCongruence}, and prove it under
the free action assumptions.
%
The coarsest congruence
contained in (weak) bisimilarity, namely \emph{bisimilarity
  congruence}, is
the \emph{composition closure} (\texttt{CC}) of (weak) bisimilarity:
\begin{alltt}
   \HOLConst{CC} \HOLFreeVar{R} \HOLTokenDefEquality{} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{g} \HOLBoundVar{h}. \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{c}. \HOLConst{CONTEXT} \HOLBoundVar{c} \HOLSymConst{\HOLTokenImp{}} \HOLFreeVar{R} \ensuremath{(}\HOLBoundVar{c} \HOLBoundVar{g}\ensuremath{)} \ensuremath{(}\HOLBoundVar{c} \HOLBoundVar{h}\ensuremath{)}\ensuremath{)}\hfill{[CC_def]}
   \HOLConst{WEAK_CONGR} \HOLTokenDefEquality{} \HOLConst{CC} \HOLConst{WEAK_EQUIV}\hfill{[WEAK_CONGR]}
\end{alltt}
We do not need to put $R\ g\ h$ \hl{into} the antecedents of
  \texttt{CC\_def}, as this is anyhow obtained from the trivial context $(\lambda x.\,x)$.
The next result shows that, for any binary relation $R$ 
on CCS processes, the composition closure of $R$ is always at least as
fine as $R$: ($\subseteq_r$ stands for \emph{relational subset}.)
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{R}. \HOLConst{CC} \HOLBoundVar{R} \HOLSymConst{\HOLTokenRSubset{}} \HOLBoundVar{R}\hfill{[CC_is_finer]}
\end{alltt}
Furthermore, we prove that any (pre)congruence contained in $R$, \hl{who
itself} needs not to be a (pre)congruence,
is contained in the composition closure of $R$
(hence the \hl{composition} closure is the coarsest one):
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{R} \ensuremath{\HOLBoundVar{R}\sp{\prime}}. \HOLConst{congruence} \ensuremath{\HOLBoundVar{R}\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \ensuremath{\HOLBoundVar{R}\sp{\prime}} \HOLSymConst{\HOLTokenRSubset{}} \HOLBoundVar{R} \HOLSymConst{\HOLTokenImp{}} \ensuremath{\HOLBoundVar{R}\sp{\prime}} \HOLSymConst{\HOLTokenRSubset{}} \HOLConst{CC} \HOLBoundVar{R}\hfill{[CC_is_coarsest]}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{R} \ensuremath{\HOLBoundVar{R}\sp{\prime}}. \HOLConst{precongruence} \ensuremath{\HOLBoundVar{R}\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \ensuremath{\HOLBoundVar{R}\sp{\prime}} \HOLSymConst{\HOLTokenRSubset{}} \HOLBoundVar{R} \HOLSymConst{\HOLTokenImp{}} \ensuremath{\HOLBoundVar{R}\sp{\prime}} \HOLSymConst{\HOLTokenRSubset{}} \HOLConst{CC} \HOLBoundVar{R}\hfill{[PCC_is_coarsest]}
\end{alltt}

Given the central role of the  
 summation, we also consider the relation closure of bisimilarity
 w.r.t. summation, called \emph{equivalence compatible with summation}
(\texttt{SUM_EQUIV}): %%, denoted by $\approx^+$: (not used this symbol)
\begin{alltt}
   \HOLConst{SUM_EQUIV} \HOLTokenDefEquality{} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{p} \HOLBoundVar{q}. \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLBoundVar{p} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{q} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r}\ensuremath{)}\hfill{[SUM_EQUIV]}
\end{alltt}

% it doesn't satisfy substitutivity on direct sums (but if the CCS syntax
% is non-standard, i.e.~has only prefixed sums, $\approx$ is indeed a
% congruence). The purpose is to find a coarsest congruence contained in
% weak bisimilarity. (``coarsest'' means, any other congruence finer than it must be contained in it)
% There're two ways to build a congruence from weak bisimilarity, one
% way is the standard definition for observational congruence (rooted
% weak bisimilarity) $\rapprox$ in CCS textbooks, but even it's proven to be a
% congruence we don't know if it's coarsest one.  The other way is to
% build a (pre)congruence closure (Def ??) directly upon
% the original weak bisimilarity relation, we call the resulting
% relation ``Weak bisimilarity congruence'' ($[\approx]$):
% \begin{alltt}
% \HOLConst{WEAK_CONGR} \HOLTokenDefEquality{} \HOLConst{CC} \HOLConst{WEAK_EQUIV}
% \end{alltt}
% It can be shown that any such (pre)congruence closure is automatically coarsest.
%
% Now it remains to prove that, the congruence relation built by above
% two quite different approaches actually coincide. To achieve this goal,
% we first noticed that, all other operators beside sums used in
% semantic context doesn't matter, because they're already substitutable
% for weak bisimilarity. The only important operator is the sum
% operator. To focus on this important operator, we can temporally
% introduce another concept called \emph{sum equivalence}:
% \begin{alltt}
% \HOLConst{SUM_EQUIV} \HOLTokenDefEquality{} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{p} \HOLBoundVar{q}. \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLBoundVar{p} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{q} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r}\ensuremath{)}
% \end{alltt}
Rooted bisimilarity $\rapprox$ (a congruence contained in
$\wb$), is now contained in \texttt{WEAK_CONGR},
which in turn is trivially contained in \texttt{SUM_EQUIV}, as shown
in Fig.~\ref{fig:relationship}. Thus, to prove \hl{Prop.}~\ref{prop:coarsest},
the crux is to prove that \texttt{SUM_EQUIV} implies
rooted bisimilarity ($\rapprox$), making all three relations
($\rapprox$, \texttt{WEAK_CONGR} and \texttt{SUM_EQUIV}) coincide:
\begin{equation}
\label{equa:pq}
\forall p\ \ q.\ ( \forall r.\ p\ +\ r \;\approx\; q\ +\ r ) \
\Longrightarrow\ p\ \rapprox\ \! q\enspace.
\end{equation}

@@

\hl{The} actual formalisation of (\ref{equa:pq}) says:
\begin{theorem}[\texttt{COARSEST_CONGR_RL}]
  Under the free actions assumption, $\rapprox$ is coarsest congruence contained in $\wb$.
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{p} \HOLBoundVar{q}. \HOLConst{free_action} \HOLBoundVar{p} \HOLSymConst{\HOLTokenConj{}} \HOLConst{free_action} \HOLBoundVar{q} \HOLSymConst{\HOLTokenImp{}} \ensuremath{(}\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLBoundVar{p} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{q} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r}\ensuremath{)} \HOLSymConst{\HOLTokenImp{}} \HOLBoundVar{p} \HOLSymConst{\HOLTokenObsCongr} \HOLBoundVar{q}
\end{alltt}
\end{theorem}

With an almost identical proof, rooted contraction
($\rcontr$) is also the coarsest
precongruence contained in the bisimilarity contraction ($\mcontrBIS$)
(the other direction is trivial):
\begin{theorem}[\texttt{COARSEST_PRECONGR_RL}]
  Under the free actions assumption, $\mcontrBIS$ is the coarsest precongruence contained in $\contr$.
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{p} \HOLBoundVar{q}. \HOLConst{free_action} \HOLBoundVar{p} \HOLSymConst{\HOLTokenConj{}} \HOLConst{free_action} \HOLBoundVar{q} \HOLSymConst{\HOLTokenImp{}} \ensuremath{(}\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLBoundVar{p} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r} \HOLSymConst{\HOLTokenContracts{}} \HOLBoundVar{q} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r}\ensuremath{)} \HOLSymConst{\HOLTokenImp{}} \HOLBoundVar{p} \HOLSymConst{\HOLTokenObsContracts} \HOLBoundVar{q}
\end{alltt}
\end{theorem}

The formal proofs  of the above two results follow Milner's proof~\cite{Mil89}.
The assumption in Milner's result 
requires $\mathrm{fn}(p) \cup
\mathrm{fn}(q) \neq \mathscr{L}$ (where $\mathrm{fn}$ stands for \emph{free
  names}), and is therefore strictly stronger than the assumption in
our result above. Indeed, in our result we only look at immediate weak
derivatives for $p$ and $q$, and only requires that there is an input
or output action
that never appears as a label. Furthermore, such an action need
not to be the same for $p$ and for $q$,
as an action $a$ and its complementary $\outC a$ are
considered as \emph{different} actions.

@@

In this section, we give another more complex proof of the second part
of Theorem~\ref{t:rapproxCongruence}, but only for finite-state CCS:
\begin{theorem}[\texttt{COARSEST_CONGR_FINITE}]
    \label{thm:coarsestfiniteState}
    For finite-state CCS, $\wb$ is the coarsest congruence contained in $\approx$:
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{p} \HOLBoundVar{q}. \HOLConst{finite_state} \HOLBoundVar{p} \HOLSymConst{\HOLTokenConj{}} \HOLConst{finite_state} \HOLBoundVar{q} \HOLSymConst{\HOLTokenImp{}} \ensuremath{(}\HOLBoundVar{p} \HOLSymConst{\HOLTokenObsCongr} \HOLBoundVar{q} \HOLSymConst{\HOLTokenEquiv{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLBoundVar{p} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{q} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r}\ensuremath{)}
\end{alltt}
\end{theorem}

>>>>

In this section we give a proof of the second part of
Theorem~\ref{t:rapproxCongruence}, i.e. $\rapprox$ is the coarsest
congruence contained in $\wb$.
The formalisation of this theorem presents some 
delicate aspects; for instance, 
with   our CCS syntax
(with only binary sums)
we have to add  an hypothesis about 
 \hl{the involved processes  not
  using all available labels}.
% This theorem is of special \hl{interest} to us, because under our
% \hl{CCS syntax (with only binary summations)} it seems impossible to
% prove it without any \hl{antecedent such as ``the involved processes
%   do not use up all available labels''}.
The general form of the theorem is thus~\cite{van2005characterisation,Gorrieri:2015jt,Mil89}:
\begin{proposition}
\label{prop:coarsest}
  \hl{Rooted bisimilarity ($\rapprox$) is the coarsest congruence
    contained in weak bisimilarity ($\wb$):}
  \begin{equation}
    \label{eq:coarsest}
\forall p\ \ q.\ p\ \rapprox\ \! q\ \Longleftrightarrow\ ( \forall r.\ p\ +\
r\ \approx\ q\ +\ r )\enspace.
\end{equation}
\end{proposition}
From left to right (\ref{eq:coarsest}) trivially holds, due to the substitutivity of
$\rapprox$ \hl{for} summation and \hl{the fact that} $\rapprox$ implies $\wb$:
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{p} \HOLBoundVar{q}. \HOLBoundVar{p} \HOLSymConst{\HOLTokenObsCongr} \HOLBoundVar{q} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLBoundVar{p} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{q} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r}\hfill{[COARSEST_CONGR_LR]}
\end{alltt}
Thus we are only interested in (\ref{eq:coarsest}) from right to left.
%
The standard argument~\cite{Mil89} requires that $p$
and $q$ do not use \hl{all} available labels (i.e.~visible actions).
Formalising this \hl{result, however,} requires a detailed treatment of free and bound names of CCS
processes (with the restriction operator acting as a binder).
\hl{In} our formalisation, in contrast, 
%  we do not discuss free and bound names (also
% called ``sorts'') of CCS processes. Instead, 
we 
assume \hl{the weaker condition} that all immediate \emph{weak} derivatives of
 $p$ and $q$ do not use all available \hl{labels}. We  call this
 the \emph{free action} property:
\begin{alltt}
   \HOLConst{free_action} \HOLFreeVar{p} \HOLTokenDefEquality{} \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{a}. \HOLSymConst{\HOLTokenForall{}}\ensuremath{\HOLBoundVar{p}\sp{\prime}}. \HOLSymConst{\HOLTokenNeg{}}\ensuremath{(}\HOLFreeVar{p} \HOLTokenWeakTransBegin\HOLConst{label} \HOLBoundVar{a}\HOLTokenWeakTransEnd \ensuremath{\HOLBoundVar{p}\sp{\prime}}\ensuremath{)}\hfill{[free_action_def]}
\end{alltt}

Below we show how \hl{Prop.}~\ref{prop:coarsest} is connected with
the statement in Theorem~\ref{t:rapproxCongruence}, and prove it under
the free action assumptions.
%
The coarsest congruence
contained in (weak) bisimilarity, namely \emph{bisimilarity
  congruence}, is
the \emph{composition closure} (\texttt{CC}) of (weak) bisimilarity:
\begin{alltt}
   \HOLConst{CC} \HOLFreeVar{R} \HOLTokenDefEquality{} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{g} \HOLBoundVar{h}. \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{c}. \HOLConst{CONTEXT} \HOLBoundVar{c} \HOLSymConst{\HOLTokenImp{}} \HOLFreeVar{R} \ensuremath{(}\HOLBoundVar{c} \HOLBoundVar{g}\ensuremath{)} \ensuremath{(}\HOLBoundVar{c} \HOLBoundVar{h}\ensuremath{)}\ensuremath{)}\hfill{[CC_def]}
   \HOLConst{WEAK_CONGR} \HOLTokenDefEquality{} \HOLConst{CC} \HOLConst{WEAK_EQUIV}\hfill{[WEAK_CONGR]}
\end{alltt}
We do not need to put $R\ g\ h$ \hl{into} the antecedents of
  \texttt{CC\_def}, as this is anyhow obtained from the trivial context $(\lambda x.\,x)$.
The next result shows that, for any binary relation $R$ 
on CCS processes, the composition closure of $R$ is always at least as
fine as $R$ ($\subseteq_r$ stands for \emph{relational subset}):
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{R}. \HOLConst{CC} \HOLBoundVar{R} \HOLSymConst{\HOLTokenRSubset{}} \HOLBoundVar{R}\hfill{[CC_is_finer]}
\end{alltt}
Furthermore, we prove that any (pre)congruence contained in $R$, \hl{that
itself} needs not to be a (pre)congruence,
is contained in the composition closure of $R$
(hence the \hl{composition} closure is the coarsest one):
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{R} \ensuremath{\HOLBoundVar{R}\sp{\prime}}. \HOLConst{congruence} \ensuremath{\HOLBoundVar{R}\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \ensuremath{\HOLBoundVar{R}\sp{\prime}} \HOLSymConst{\HOLTokenRSubset{}} \HOLBoundVar{R} \HOLSymConst{\HOLTokenImp{}} \ensuremath{\HOLBoundVar{R}\sp{\prime}} \HOLSymConst{\HOLTokenRSubset{}} \HOLConst{CC} \HOLBoundVar{R}\hfill{[CC_is_coarsest]}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{R} \ensuremath{\HOLBoundVar{R}\sp{\prime}}. \HOLConst{precongruence} \ensuremath{\HOLBoundVar{R}\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \ensuremath{\HOLBoundVar{R}\sp{\prime}} \HOLSymConst{\HOLTokenRSubset{}} \HOLBoundVar{R} \HOLSymConst{\HOLTokenImp{}} \ensuremath{\HOLBoundVar{R}\sp{\prime}} \HOLSymConst{\HOLTokenRSubset{}} \HOLConst{CC} \HOLBoundVar{R}\hfill{[PCC_is_coarsest]}
\end{alltt}

Given the central role of  
 summation, we also consider the relation closure of bisimilarity
 w.r.t. summation, called \emph{equivalence compatible with summation}
(\texttt{SUM_EQUIV}): %%, denoted by $\approx^+$: (not used this symbol)
\begin{alltt}
   \HOLConst{SUM_EQUIV} \HOLTokenDefEquality{} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{p} \HOLBoundVar{q}. \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLBoundVar{p} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{q} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r}\ensuremath{)}\hfill{[SUM_EQUIV]}
\end{alltt}

% it doesn't satisfy substitutivity on direct sums (but if the CCS syntax
% is non-standard, i.e.~has only prefixed sums, $\approx$ is indeed a
% congruence). The purpose is to find a coarsest congruence contained in
% weak bisimilarity. (``coarsest'' means, any other congruence finer than it must be contained in it)
% There're two ways to build a congruence from weak bisimilarity, one
% way is the standard definition for observational congruence (rooted
% weak bisimilarity) $\rapprox$ in CCS textbooks, but even it's proven to be a
% congruence we don't know if it's coarsest one.  The other way is to
% build a (pre)congruence closure (Def ??) directly upon
% the original weak bisimilarity relation, we call the resulting
% relation ``Weak bisimilarity congruence'' ($[\approx]$):
% \begin{alltt}
% \HOLConst{WEAK_CONGR} \HOLTokenDefEquality{} \HOLConst{CC} \HOLConst{WEAK_EQUIV}
% \end{alltt}
% It can be shown that any such (pre)congruence closure is automatically coarsest.
%
% Now it remains to prove that, the congruence relation built by above
% two quite different approaches actually coincide. To achieve this goal,
% we first noticed that, all other operators beside sums used in
% semantic context doesn't matter, because they're already substitutable
% for weak bisimilarity. The only important operator is the sum
% operator. To focus on this important operator, we can temporally
% introduce another concept called \emph{sum equivalence}:
% \begin{alltt}
% \HOLConst{SUM_EQUIV} \HOLTokenDefEquality{} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{p} \HOLBoundVar{q}. \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLBoundVar{p} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{q} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r}\ensuremath{)}
% \end{alltt}
Rooted bisimilarity $\rapprox$ (as a congruence contained in
$\wb$) is now contained in \texttt{WEAK_CONGR},
which in turn is trivially contained in \texttt{SUM_EQUIV}, as shown
in Fig.~\ref{fig:relationship}. Thus, to prove \hl{Prop.}~\ref{prop:coarsest},
the crux is to prove that \texttt{SUM_EQUIV} implies
rooted bisimilarity ($\rapprox$), making all three relations
($\rapprox$, \texttt{WEAK_CONGR} and \texttt{SUM_EQUIV}) coincide:
\begin{equation}
\label{equa:pq}
\forall p\ \ q.\ ( \forall r.\ p\ +\ r \;\approx\; q\ +\ r ) \
\Longrightarrow\ p\ \rapprox\ \! q\enspace.
\end{equation}

@@

\hl{The}  formalisation of (\ref{equa:pq}) says:
\begin{theorem}[\texttt{COARSEST_CONGR_RL}]
  Under the free actions assumption, $\rapprox$ is coarsest congruence contained in $\wb$.
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{p} \HOLBoundVar{q}. \HOLConst{free_action} \HOLBoundVar{p} \HOLSymConst{\HOLTokenConj{}} \HOLConst{free_action} \HOLBoundVar{q} \HOLSymConst{\HOLTokenImp{}} \ensuremath{(}\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLBoundVar{p} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{q} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r}\ensuremath{)} \HOLSymConst{\HOLTokenImp{}} \HOLBoundVar{p} \HOLSymConst{\HOLTokenObsCongr} \HOLBoundVar{q}
\end{alltt}
\end{theorem}

With an almost identical proof, rooted contraction
($\rcontr$) is also the coarsest
precongruence contained in the bisimilarity contraction ($\mcontrBIS$):
\begin{theorem}[\texttt{COARSEST_PRECONGR_RL}]
  Under the free actions assumption, $\mcontrBIS$ is the coarsest precongruence contained in $\contr$.
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{p} \HOLBoundVar{q}. \HOLConst{free_action} \HOLBoundVar{p} \HOLSymConst{\HOLTokenConj{}} \HOLConst{free_action} \HOLBoundVar{q} \HOLSymConst{\HOLTokenImp{}} \ensuremath{(}\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLBoundVar{p} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r} \HOLSymConst{\HOLTokenContracts{}} \HOLBoundVar{q} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r}\ensuremath{)} \HOLSymConst{\HOLTokenImp{}} \HOLBoundVar{p} \HOLSymConst{\HOLTokenObsContracts} \HOLBoundVar{q}
\end{alltt}
\end{theorem}

The formal proofs  of the above two results follow Milner's proof~\cite{Mil89}.
The assumption in Milner's result 
requires $\mathrm{fn}(p) \cup
\mathrm{fn}(q) \neq \mathscr{L}$ (where $\mathrm{fn}$ stands for \emph{free
  names}), and is therefore strictly stronger than the assumption in
our result above. Indeed, in our result we only look at immediate weak
derivatives for $p$ and $q$, and only require that there is an input
or output action
that never appears as a label. Furthermore, such an action need
not to be the same for $p$ and for $q$,
and an action $a$ and its complementary $\outC a$ are
considered {different} actions.

@@

In this section, we give another more complex proof of the second part
of Theorem~\ref{t:rapproxCongruence}, but only for finite-state CCS.
\begin{theorem}[\texttt{COARSEST_CONGR_FINITE}]
    \label{thm:coarsestfiniteState}
    For finite-state CCS, $\wb$ is the coarsest congruence contained in $\approx$:
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{p} \HOLBoundVar{q}. \HOLConst{finite_state} \HOLBoundVar{p} \HOLSymConst{\HOLTokenConj{}} \HOLConst{finite_state} \HOLBoundVar{q} \HOLSymConst{\HOLTokenImp{}} \ensuremath{(}\HOLBoundVar{p} \HOLSymConst{\HOLTokenObsCongr} \HOLBoundVar{q} \HOLSymConst{\HOLTokenEquiv{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLBoundVar{p} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{q} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r}\ensuremath{)}
\end{alltt}
\end{theorem}

The formal proof follows van
Glabbeek~\cite{van2005characterisation}.
The core lemma says that for
any two processes $p$ and $q$, if there exists a \emph{stable} (i.e.~$\tau$-free)
 process $k$ which is not bisimilar with any derivative of $p$ and
 $q$, then \texttt{SUM_EQUIV} indeed implies rooted bisimilarity
 ($\rapprox$) (c.f.~\cite{Tian:2017wrba} for more details of the proof):
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{p} \HOLBoundVar{q}.
       \ensuremath{(}\HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{k}.
            \HOLConst{STABLE} \HOLBoundVar{k} \HOLSymConst{\HOLTokenConj{}} \ensuremath{(}\HOLSymConst{\HOLTokenForall{}}\ensuremath{\HOLBoundVar{p}\sp{\prime}} \HOLBoundVar{u}. \HOLBoundVar{p} \HOLTokenWeakTransBegin\HOLBoundVar{u}\HOLTokenWeakTransEnd \ensuremath{\HOLBoundVar{p}\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}\ensuremath{(}\ensuremath{\HOLBoundVar{p}\sp{\prime}} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{k}\ensuremath{)}\ensuremath{)} \HOLSymConst{\HOLTokenConj{}}
            \HOLSymConst{\HOLTokenForall{}}\ensuremath{\HOLBoundVar{q}\sp{\prime}} \HOLBoundVar{u}. \HOLBoundVar{q} \HOLTokenWeakTransBegin\HOLBoundVar{u}\HOLTokenWeakTransEnd \ensuremath{\HOLBoundVar{q}\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}\ensuremath{(}\ensuremath{\HOLBoundVar{q}\sp{\prime}} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{k}\ensuremath{)}\ensuremath{)} \HOLSymConst{\HOLTokenImp{}}
       \ensuremath{(}\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLBoundVar{p} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{q} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r}\ensuremath{)} \HOLSymConst{\HOLTokenImp{}}
       \HOLBoundVar{p} \HOLSymConst{\HOLTokenObsCongr} \HOLBoundVar{q}\hfill{[PROP3_COMMON]}
\end{alltt}
\begin{alltt}
   \HOLConst{STABLE} \HOLFreeVar{p} \HOLTokenDefEquality{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{u} \ensuremath{\HOLBoundVar{p}\sp{\prime}}. \HOLFreeVar{p} \HOLTokenTransBegin\HOLBoundVar{u}\HOLTokenTransEnd \ensuremath{\HOLBoundVar{p}\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLBoundVar{u} \HOLSymConst{\HOLTokenNotEqual{}} \HOLSymConst{\ensuremath{\tau}}\hfill{[STABLE]}
\end{alltt}

To actually construct this process $k$,
 the proof relies on arbitrary infinite sums of 
processes and uses transfinite induction to obtain
an arbitrary large sequence of processes (firstly introduced by Jan
Willem Klop \cite{van2005characterisation})
 that are all pairwise non-bisimilar.
We have only partially formalised
this proof (Theorem~\ref{thm:coarsestfiniteState} in next section),
because
the typed logic
implemented in various HOL systems (including Isabelle/HOL) is not
strong enough to define a type for all possible
ordinals~\cite{norrish2013ordinals}.
As a consequence, the final result is only for finite-state
CCS.\footnote{\hl{van Glabbeek's
  original proof applies to all CCS processes; however,  it
  requires the CCS syntax to support arbitrary large summations based
  on ordinals, and relies on the existence  of an
  ordinal larger than the cardinality of any given set. This is beyond
  the possibilities of HOL.}}
